---
title: 'STA304'
author: "Neil Montgomery"
date: "2016-01-28"
output: 
  ioslides_presentation: 
    css: 'stylesOI.css' 
    widescreen: true 
    transition: 0.001
---
\newcommand{\E}[1]{E{\left(#1\right)}}
\newcommand{\flist}[2]{\{#1_1, #1_2, \ldots, #1_#2\}}
\renewcommand{\bar}[1]{\overline{#1}}
\newcommand{\SE}[1]{\sqrt{\hat{V}(#1)}}

```{r, echo=FALSE}
options(scipen = 999, digits=3)
```


# proportions (and counts)

## categorical data 

A significant amount of sample data is categorical in nature, as opposed to fundamentally numerical. But the line can be blurry.

* "Failed" vs. "Not Failed"
* "Liberal", "Conservative", "NDP", "Green", "BQ"
* Gas pipes come in 1.0, 1.5, and 1.75 inch sizes
* Likert scale questions (thousands of papers arguing about these!)

In this course we'll only consider the case of two outcomes, which is not as limiting as it sounds.

LifeProTip: use *proportions* and not *percentages*---the latter are merely formatting conventions for human viewing.

## analogy/notation

Recall the actual population:
$$\flist{e}{N}$$
which we have usually conflated with (numerical) *measurements taken* on the 
population:
$$\flist{y}{N}$$

For the case of proportions and counts, we can simply (arbitrarily) code the two outcomes as 0 and 1 and then use all the previous theory.

But just to mess with you we'll change the notation. Rather than $\mu$ (the population mean) we'll use $p$ (the population proportion of 1s) although the definition is identical: $p = \frac{1}{N}\sum_{i=1}^N y_i$. For convenience we'll also define $q = 1 - p$ (the population proportion of 0s)

## the population variance (not in book!) { .build }

The population variance is *defined* (like before) as: $\sigma^2 = \frac{1}{N}\sum_{i=1}^N \left(y_i - p\right)^2$. This turns out to be:

$$\begin{align*}
\sigma^2=\frac{1}{N}\sum_{i=1}^N \left(y_i - p\right)^2 &= \frac{1}{N}\left(\sum_{i=1}^N y_i^2 - Np^2\right)\\
&= \frac{1}{N}\left(\sum_{i=1}^N y_i - Np^2\right)\\
&= \frac{1}{N}\left(Np - Np^2\right)\\
&= p(1-p) = pq
\end{align*}$$

## the sample and its properties

So we don't know what $p$ is. We plan to gather a sample $\flist{y}{n}$. Under simple random sample this sample has some familiar properties.

>* The $y_i$ have the same distribution with pmf as follows
$$\begin{equation*}
y_i = \begin{cases}
1 & \text{with probability } p,\\
0 & \text{with probability } 1-p = q
\end{cases}
\end{equation*}$$
*(stats people would prefer the more useful: $p(y) = p^y(1-p)^{1-y}$ for $y\in\{0,1\}$)*

>* $\E{y_i} = p$ and $V(y_i) = pq$

>* But the $y_i$ are *not* independent...

>* We won't use this, but what is the *distribution* of $k$, the number of 1s (some may know)?

## estimating $p$ under SRS { .build }

Use the usual estimator with a new name:
$$\bar{y} = \hat{p} = \frac{1}{n}\sum_{i=1}^n y_i$$

If I need to I might also declare $k$ to be the "number of 1s" in the sample and use $\hat{p} = k/n$.

For convenience we'll also define $\hat{q} = 1 - \hat{p}$.

$\hat{p}$ is unbiased for $p$ (easy exercise)

## the true variance of $\hat{p}$ { .build }

Since there's nothing special about 0s and 1s for population values we can recycle the old methods:
$$\begin{align*}
V(\hat{p}) &= \frac{\sigma^2}{n}\left(\frac{N-n}{N-1}\right) \text{ (copied from before)}\\
&= \frac{pq}{n}\left(\frac{N-n}{N-1}\right) \text{ (since $\sigma^2 = pq$)}
\end{align*}$$

And, $s^2 = \frac{1}{n-1}\sum_{i=1}^n \left(y_i - \hat{p}\right)^2$ is still used to estimate $\sigma^2 = pq$. It turns out that $s^2 = \frac{n\hat{p}\hat{q}}{n-1}$, so we end up with this unbiased estimator for $V(\hat{p})$:
$$\hat{V}(\hat{p}) = \frac{\hat{p}\hat{q}}{n-1}\frac{N-n}{N}$$
Due to symmetry $\hat{V}(\hat{q}) = \hat{V}(\hat{p})$.

## (counts)

Population total remains: $\tau = \sum_{i=1}^N y_i$. No special notation this time.

Results: use $Np=\tau$, so essentially multiply by $N$ as appropriate. 

## example { .build }

A gas distribution company has employees and contractors who install and repair equipment. It is important that the work be performed up to standard. The company has a Quality Assurance (QA) program that audits samples of work done. The QA program selects a SRS of $n=`r (n <- 412)`$ out of $N=`r (N <- 12251)`$ tasks to audit. A task "conforms" if it meets all "technical" *and* "paperwork" standards.

In the SRS $k = `r (k <- 377)`$ tasks were found to conform. Estimate the proportion of conforming tasks. Estimate the number of non-conforming tasks that will require follow-up work.

We have $\hat{p} = `r (p <- k/n)`$ with standard error $\SE{\hat{p}} = \sqrt{\frac{\hat{p}\hat{q}}{n-1}\frac{N-n}{N}} = `r (se <- sqrt(p*(1-p)/(n-1)*(N-n)/N))`$ and a 95% C.I. is $[`r p-2*se`, `r p+2*se`]$.

A 95% C.I. for the number of non-conforming tasks (simply $Nq$, estimated by $N\hat{q}$) is then $[`r N*((1-p)-2*se)`, `r N*((1-p)+2*se)`]$.

## sample size selection for proportions { .build }

For a bound $B$ on estimation error and assuming 95% confidence level, the recycled formula becomes:
$$n = \frac{N\sigma^2}{(N-1)B^2/4 + \sigma^2} = \frac{Npq}{(N-1)B^2/4 + pq}$$
along with the classic "but we don't know" (...in this case...) "$p$" problem.

<div class="columns-2">
```{r, echo=FALSE, message=FALSE, fig.width=4, fig.height=3}
library(dplyr)
library(ggplot2)
pdata <- data.frame(p = seq(0, 1, length.out = 1000))
pqdata <- mutate(pdata, pq=p*(1-p))
pqdata %>% 
  ggplot(aes(p, pq)) + geom_line()
```

How to handle this problem requires an understanding of $pq=p(1-p)$ as a function of $p$. 

Maximum at $p=0.5$ and symmetric. 
</div>

## handling the unknown $p$ problem 

The basic rule is: use as a "guess" the closest value to 0.5 that prior knowledge gives you. Examples:

>* You have good information that the true value is between 0.2 and 0.3. **Use 0.3 in the sample size formula**.

>* You have good information that the true value is between 0.75 and 0.85. **Use 0.75 in the sample size formula**

>* You have no information. **Use 0.5**.

>* You have good information that the true value is between 0.4 and 0.6. **Use 0.5**.

## example { .build }

For the next audit cycle the QA department wants a closer estimate of the amount of follow-up work that will have to be done. Last time the estimate was within $N\SE{\hat{q}} = `r 2*N*se`$. This time they want to be within 200. What should the sample size be this time?

The sample size formula $n = \frac{Npq}{(N-1)B^2/4 + pq}$ is in terms of a bound $B$ for the error in estimating $p$ (or $q$). We're interested in the total $Nq$. So in the formula put $B/N$ in place of $B$.

There's also the matter of the unknown $p$. We can use the 95% C.I. from before as our "good information": $[`r p-2*se`, `r p+2*se`]$. The closest to 0.5 is $`r (pm <- p - 2*se)`$.

## example - answer

So the formula becomes:

$$\begin{align*}
n &= \frac{Npq}{(N-1)(B/N)^2/4 + pq}\\
&= \frac{`r N`(`r pm`)(`r 1-pm`)}{(`r N-1`)(`r sprintf("%8f", (200/12251)^2/4)`)+`r pm`(`r (1-pm)`)}\\
&= `r N*pm*(1-pm)/((N-1)*(200/N)^2/4 + pm*(1-pm))`
\end{align*}$$

## caution: rare items { .build }

The gas company adopts a policy to estimate proportions with a bound of
$B = `r (B <- 0.05)`$ on the error of estation (95% confidence level assumed).

Suppose the gas company operates $N= `r (N <- 10000)`$ of a certain type of valve. They want to estimate the proportion that are "stuck" open (which is very dangerous). 

So they use the sample size formula with the maximal $p=`r (pm <- 0.5)`$ to obtain a required sample size of $n = `r N*pm*(1-pm)/((N-1)*(B)^2/4 + pm*(1-pm))`$.

They collect a sample of that size and find $k= `r (k <- 1)`$ (!) stuck valve in the sample. 

```{r, echo=FALSE}
options(digits=6)
p <- k/n
se <- sqrt(p*(1-p)/n*(N-n)/N)
```

With some concern they calculate a confidence interval and get $[`r p - 2*se`, `r p + 2*se`]$. They contact a statistician for assistance. 


